---
title: "Inference Models"
---

This page features information about creating and interpreting regression models.

- Simple Linear Regression (math, programming, and interpretation)
- Multiple Linear Regression (math, programming, and interpretation)

# Simple Linear Regression

The goal of simple linear regression is to figure out an equation of a line through the data in an attempt to find the relationship between feature and response to predict future values. The equation is of the form:

$y_i = \alpha + \beta x_i + \epsilon_i$

where

- $y_i$ is the prediction at $x_i$
- $\alpha$ is the "y-intercept"
- $\beta$ is the "slope"
- $\epsilon_i \sim N(0, \sigma^2)$, and each are independent

In vector (or set) form:

$Y = \alpha + \beta X + \epsilon$

where

- $X$: the independent variable, the predictor, the explanatory variable, the feature
- $Y$: the dependent variable, the response variable
- $\epsilon$: the random deviation, random error – accounts for the fact that the world is uncertain and that there are random deviations around the true process.

## The Expected Value of the Response Variable

Given $Y = \alpha + \beta X + \epsilon$, what is $E[Y]$?

$E[Y] = E[\alpha + \beta X + \epsilon]$

$= E[\alpha] + E[\beta X] + E[\epsilon]$

$= \alpha + \beta E[X] + 0$

$= \alpha + \beta X$

## Interpreting Parameters

$\beta$: the slope of our true regression line represents the increase in our response from a unit increase in our feature.

## Minimizing the Residuals

Given our data, $(x_1, y_1), \dots, (x_n, y_n)$ how do we minimize the residuals, $\epsilon_i = y_i - (\alpha + \beta x_i)$? In other words, how do we minimize the sum of the squared errors?

> insert image of vertical line from regression line to point, label it e_i

Given:

- $y_i$: the actual value of the data point
- $\hat{y_i}$: the predicted value of the $i^{th}$ data point

$SSE = \sum\limits_{i=1}^{n} (y_i - \hat{y_i})^2$

> the point-estimates (single value estimated from the data) of the slope and intercept parameters are called the least-squares estimates, and are defined to be the values that minimize the SSE. The SSE can be thought of as a measure of how much variation in $Y$ is left unexplained by the model.

## How we Find the Parameter Estimates?

$SSE = \sum\limits_{i=1}^{n} (y_i - \hat{y_i})^2$

$= \sum\limits_{i=1}^{n} (y_i - (\alpha - \beta x_i))^2$

**After making this substitution, compute...**

- $\frac{\partial SSE}{\partial \alpha} = 0$
- $\frac{\partial SSE}{\partial \beta} = 0$

**Which yield the following solutions, respectively...**

- $\alpha = \bar{y} - \beta \bar{x}$
- $\beta = \frac{\bar{xy} - \bar{x} \bar{y}}{\bar{x^2} - (\bar{x})^2}$

## Estimating the Variance

Given that the parameter $\sigma^2$ determines the spread of the data about the true regression line, our estimate of the variance is $\hat{\sigma^2}$, which is equivalent to:

$\hat{\sigma^2} = \frac{SSE}{n-2}$

## Coefficient of Determination

The coefficient of determination, $R^2$, quantifies how well the model explains the data. $R^2$ ranges from $0$ to $1$.

The regression sum of squares is given by:

$SSR = \sum\limits_{i=1}^n (\hat{y_i} - \bar{y})^2$

and gives a sense of how much variation in $Y$ is explained by our model.

A quantitative measure of the total amount of variation in observed $Y$ values is given by the total sum of squares:

$SST = \sum\limits_{i=1}^n (y_i - \bar{y})^2$

SST is what we would get if we used the mean of the data as our model.

Note that the sum of squared deviations about the least-squares line is smaller than the sum of squared deviations about nay other line.

$SSE \leq SST$

> Include image of the three measures. 

**$R^2$**

Therefore, the ratio of $\frac{SSE}{SST}$ is the proportion of the total variation in the data (SST) that cannot be explained by the SLR model (SSE). So we define the coefficient of determination $R^2$ to be the proportion of variance that can be explained by the model.

$R^2$ has a magnitude between $0$ and $1$, with the closer to $1$ be the better (i.e. the higher the number the more of the variation that can be explained by the model).

$R^2 = 1 - \frac{SSE}{SST}$

## Slope Distribution

**Distribution**

$\hat{\beta} \sim N(\beta, \frac{\sigma^2}{\sum_{i=1}^n (x_i - \bar{x})^2}$

$SE(\hat{\beta}) = \frac{\sigma^2}{\sqrt{\sum_{i=1}^n (x_i - \bar{x})^2}}$

**Hypothesis Tests**

$H_0$: $\beta = 0$

$H_A$: $\beta \neq 0$

**Confidence Intervals**

$\hat{\beta} \pm t_{\alpha / 2, df = n-2} SE(\hat{\beta})$

## SUMMARY OF SLR

### Main Formula

$Y = \alpha + \beta X + \epsilon$

### Variance & Slope Distribution ($\beta$s)

$\hat{\sigma^2} = \frac{SSE}{n-2}$

$SE(\hat{\beta}) = \frac{\sigma^2}{\sqrt{\sum_{i=1}^n (x_i - \bar{x})^2}}$

### CI for t-tests on $\beta$

$\hat{\beta} \pm t_{\alpha / 2, df = n-2} SE(\hat{\beta})$

### Coefficient of Determination

| | SSE | SSR | SST | $R^2$ |
|---|---|---|---|---|
| Description | Measure of how much variation in $Y$ is left unexplained by the model | How much variation in $Y$ is explained by our model | Total amount of variation in observed $Y$ values (what we would get if we used the mean of the data as our model) | The proportion of variance that can be explained by the model |
| Formula | $\sum\limits_{i=1}^{n} (y_i - \hat{y_i})^2$ | $\sum\limits_{i=1}^n (\hat{y_i} - \bar{y})^2$ | $\sum\limits_{i=1}^n (y_i - \bar{y})^2$ | $1 - \frac{SSE}{SST}$ |

: {.striped .hover}

**The closer to 1 $R^2$ is, the better the model fits the data.**

## Workflow for Simple Linear Regression

1. Plot the data as a scatter plot
  a. Does linearity seem appropriate?
  b. Compute and overlay best-fit line
2. Consider assumptions in SLR
  a. Plot a histogram of the residuals (are they normal?): **WANT NORMAL** 
  b. Plot the residuals against x (are they changing?): **WANT RANDOM**
  
> Include Plots

## Simple Linear Regression in R

The following functions are paramount in implementing and interpreting simple linear regression in R:

- `lm()`
- `summary()`

> Load Library

```{r message = FALSE}
# import libraries
library(tidyverse)
```

> Load Data (use txhousing from built in data)

```{r}
df <- txhousing
head(df)
```

> Create a simple linear regression model to see if we can explain and predict volume from listings

```{r}
slr <- lm(volume ~ listings, data = df)
summary(slr)
```
From this summary, we can see that `listings` is a statistically significant factor (low p-value) and that the $R^2$ for the model is somewhat acceptable at $0.7411$. Although the intercept being negative does raise some concern, this would mean that for a time period with $0$ `listings`, we would get a negative `volume`. The model also says that for each new `listing`, we can expect the `volume` to increase by $36990$.

# Multiple Linear Regression

We can extend the **simple linear regression model** to accommodate more variables. *Note that we'll now refer to $\alpha$ as $\beta_0$*.

The main vector model can be represented as:

$Y = \beta_0 + B_1 X_1 + \dots + B_p X_p + \epsilon$

where for each of the $n$ data points (each vector has the same number of data points), we assume for $i = 1, \dots, n$, the $i^{th}$ response variable is represented as:

$y_i = \beta_0 + B_1 x_{i, 1} + \dots + B_p x_{i, p} + \epsilon_i$

## Matrix Algebra Applications & Solutions

If we have a perfectly square $X$ matrix, then we can solve for $\beta$ via:

$Y = X \beta \rightarrow X^{-1} Y = \beta$.

However, we can almost **guarantee** that $X$ will not be square. Look at this following linear algebra application:

> Given a matrix $A$ of any dimensions, both $A^TA$ and $AA^T$ will result in square matrices.

Therefore, given $Y = X \beta$, we can solve via the following:

$(X^TX)^{-1}X^TY = \beta$

## Interpreting MLR

Given $y_i = \beta_0 + B_1 x_{i, 1} + \dots + B_p x_{i, p} + \epsilon_i$, parameter $\beta_k$ is the expected change in the response associated with a unit change in the value of feature $x_k$ while all of the other features are held fixed.

## Quantifying Goodness of Fit

### $R^2$ vs. $R_a^2$

Like in SLR, we can also calculate measures like **SSE**, **SST**, and **$R^2$** for MLR.

- $SSE = \sum\limits_{i=1}^n (y_i - \hat{y_i})^2 = \sum\limits_{i=1}^n (y_i - (\beta_0 + B_1 x_{i, 1} + \dots + B_p x_{i, p})})^2$
- $SST = \sum\limits_{i=1}^n (y_i - \bar{y})^2$
- $R^2 = 1 - \frac{SSE}{SST}$

Although we can calculate $R^2$ for an MLR model, we run into the issue of multiple comparisons. The **Adjusted $R^2$**, $R_a^2$, is a better indicator of goodness of fit for MLR models. The adjusted version penalizes for having too many features that are not reducing SSE.

$R_a^2 = 1 - \frac{SSE/(n-p-1)}{SST/(n-1)}$

### Covariance & Correlation

We can discover relationships among features by performing a correlation analysis. If the value of one feature changes, how will this affect the other features.

Traditionally, these measurements are calculated via:

- Covariance: $Cov(X, Y) = E[(X - E[X])(Y -E[Y])]$
- Correlation (Pearson's): $\rho(X, Y) = \frac{Cov(X, Y)}{\sqrt{Var(X) Var(Y)}}$

However, in an MLR, we estimate these relationships using formulas analogous to the sample variance.

- Sample Covariance: $S^2_{XY} = \frac{1}{n-1} \sum\limits_{i=1}^n (x_i - \bar{x})(y_i - \bar{y})$
- Sample Correlation $\hat{\rho}(X < Y) = \frac{S^2_{XY}}{\sqrt{S^2_{X}S^2_{Y}}}$

*where $S^2_{X}$ and $S^2_{Y}$ are the variances for $X$ and $Y$, respectively*.

## Individual t-tests


















